State of the Practice of Lattice Boltzmann Method Software Development

### 2021-06-23 StateOfPractice/Peter-Notes/Report/Report.tex
LBM 5.4 feedback changes

Several development process obstacles were identified. One developer noted that their small development team has a lack of time to implement new features. Small development teams are common for LBM software packages. Team members are almost always part of the same institute or already know each other from other projects. External contributions are rare despite many of the projects accepting them.

A lack of software development experience was noted by another developer. Many of the team members on their project are domain experts and there is a steep learning curve to get these team members to contribute good quality source code. It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions. The same developer also noted that there is a lack of incentive and funding in academia for developing widely used scientific software. There are no journals that publish such source code. There is no organization helping with the development of good quality software. Scientific software is often developed in-house by the very researchers that temporarily use it in their own research. This software has a defined user and purpose, and often does not meet the standards that would be required by external users. Such software is rarely commercialized, in which case it is not open-source.

Documentation of LBM software is noted as being important, but its quality could be improved. There is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team. Furthermore, the documentation generally only provides a shallow overview of the underlying CFD theory. Users would be well advised to already be familiar with these topics to or they should spend significant time referencing theory resources. The documentation instead generally focuses on explaining how to use the software. It is of course not feasible for package documentation to address the underlying metaphysics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory. Sometimes, frequently asked questions about the underlying theory are answered in the documentation.

Setting up parallelization was also noted as a technical pain point by one of the developers, and the introduction of continuous integration by another. Software development knowledge could mitigate such pain points. As already noted, many of the developers are domain experts and not professional software developers. One of the interviewed developers that has relatively more extensive software development experience noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with reusing code to run on both the CPU and GPU.

Difficulties with ensuring correctness were also noted by several developers. They indicated that tests are run on all new source code additions, testing both individual modules and the system to verify correctness. These tests compare the package output to known correct output using test cases. A developer commented that the amount of testing data that is needed for some cases is a problem as free testing services do not offer capabilities to store and process such large amounts of data, and in-house testing solutions needed to be created to address this limitation.

A few obstacles related to users were found. Several developers noted that users sometimes try to use incorrect LBM method combinations to solve their problems. Furthermore, some users think that the packages will work out of the box to solve their cases, while the packages require both a good understanding of CFD and an understanding of the requirements for formulating problems in the individual packages, which can be a significant endeavor. These software packages are not like commercial software packages. They are generally set up to solve specific research problems, and are otfen primarily used by their developers. While they are modifiable to solve similar problems, these modifications are not trivial. So far this problem of on-boarding new users has been addressed by updating the documentation to better inform users of the underlying LBM theory and package requirements. Similar issues with LBM parameters were noted by another developer. Updating the user interface to better explain theoretical principles, as well as test user input for compatibility, was the implemented solution. As noted above, sometimes frequently asked questions on the underlying theory and on how to use the software are answered in the documentation.

One of the interviewed developers commented that parts of their package's source code had been refactored to Python to help address usability issues. Python was perceived as a much more usable language, and it would be easy for future users and developers to learn and understand the source code.

A few potential future obstacles were noted. One developer said that their source code had been written with a specific application in mind and that due to this there was too much coupling between components of the source code. This would have an impact on future modifiability and reusability when trying to extend the software, and the code would need to be refactored.

@@

>>>>

Several development process obstacles were identified. A developer of pyLBM noted that their small development team has a lack of time to implement new features. Small development teams are common for LBM software packages. Team members are almost always part of the same institute or already know each other from other projects. External contributions are rare despite many of the projects accepting them.

A lack of software development experience was noted by the developer of TCLB. Many of the team members on their project are domain experts and there can be a steep learning curve before these team members contribute good quality source code. It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions. The same developer also noted that there is a lack of incentive and funding in academia for developing widely used scientific software. The importance of funding for scientific software has been discussed in \cite{gewaltig2012quality}, which notes that ``software tools are developed and maintained only for as long as there is explicit or implicit funding''. The develop also commented that there are no journals that publish such scientific software source code. However, some work has been done to address such concerns in \citep{smith2016software}, which presents a set of software citation principles and discusses ``how they could be used to implement software citation in the scholarly community'' \citep{katz2019software}. Another raised concern was that there are no organizations helping with the development of good quality software; but some do exist, including \href{https://bssw.io/}{Better Scientific Software (BSSw)}, \href{https://www.software.ac.uk/}{Software Sustainability Institute}, and \href{https://software-carpentry.org/}{Software Carpentry}. Some SCS developers may not be familiar with these organizations.

Scientific software is often developed in-house by the very researchers that temporarily use it in their own research. Empirical studies of such ``professional end-user development'' of SCS is noted in \citep{segal2007end}. This kind of software has a defined user and purpose, and often does not meet the standards that would be required by external users. It has been categorized as a ``private tool'' by \citep{gewaltig2012quality}, which notes that despite often being made freely available, ``it is not always clear that it is sufficiently mature in terms of domain coverage, validity, documentation or usability, to be useful to other researchers''. It is less common for such ad hoc scientific software to become ``user-ready software'', which ``is not only research-ready, but should have most of the attributes commonly expected of commercial software products including broadness of scope, robustness, demonstrable correctness and adequate documentation'' \citep{gewaltig2012quality}. As software becomes user-ready, it can become commercialized and closed-source.

Documentation of LBM software is noted as being important, but its quality could be improved. There is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team. Furthermore, the documentation generally only provides a shallow overview of the underlying CFD theory. Users would be well advised to already be familiar with these topics, or they should spend significant time referencing theory resources. The documentation instead generally focuses on explaining how to use the software. It is of course not feasible for package documentation to address the underlying physics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory. Sometimes, frequently asked questions about the underlying theory are answered in the documentation. OpenLB has an artifact for such questions.

Setting up parallelization was also noted as a technical pain point by one of the developers, and the introduction of continuous integration by another. Software development knowledge, and automatic code generation, could mitigate such pain points. As already noted, many of the developers are domain experts and not professional software developers. The developer of TCLB noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with reusing code to run on both the CPU and GPU.

Difficulties with ensuring correctness were also noted by several developers. They indicated that tests are run on all new source code additions, testing both individual modules and the system to verify correctness. These tests compare the package output to known correct output using test cases. The developer of TCLB commented that the amount of testing data that is needed for some cases is a problem as free testing services do not offer capabilities to store and process such large amounts of data, and in-house testing solutions needed to be created to address this limitation. The solution for this has been to limit the size of the testing problems, and running tests in small batches of few iterations.

A few obstacles related to users were found. Several developers noted that users sometimes try to use incorrect LBM method combinations to solve their problems. Furthermore, some users think that the packages will work out of the box to solve their cases, while the packages require both a good understanding of CFD and an understanding of the requirements for formulating problems in the individual packages, which can be a significant endeavor. These software packages are not like commercial software packages. They are generally set up to solve specific research problems, and are often primarily used by their developers. While they are modifiable to solve similar problems, these modifications are not trivial. Better documentation, with attention on traceability, and automatic code generation are suggested when designing software for change, and would help with these modifications. So far this problem of on-boarding new users has been addressed by updating the documentation to better inform users of the underlying LBM theory and package requirements. Similar issues with LBM parameters were noted by another developer. Updating the user interface to better explain theoretical principles, as well as test user input for compatibility, was the implemented solution. As noted above, sometimes frequently asked questions on the underlying theory and on how to use the software are answered in the documentation.

The interviewed developer of ESPResSo commented that parts of their package's source code had been refactored to Python to help address usability issues. Python was perceived as a much more usable language, and it would be easy for future users and developers to learn and understand the source code.

A few potential future obstacles were noted. The developer of ESPResSo noted that their source code had been written with a specific application in mind and that due to this there was too much coupling between components of the source code. This results in technical debt, and would have an impact on future modifiability and reusability when trying to extend the software, and the code would need to be refactored.

@@

The developer comments have emphasized an importance on source code, while documentation seems to be of secondary importance. It must be stressed that improved management of documentation could improve package source code and help eliminate some of the developer concerns that were raised.
